---
title: "Reading Proficiency by Grade Level"
description: |
  Reading Profiency by Grade Level, Comparing Economically Disadvantaged Students to All Students.
author:
  - name: Aden Bhagwat, Emily Mingus, Erick Njue
    url: https://adenb0220.github.io/EDLD_652_Final/
date: 2025-03-01
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}

library(rio)
library(here)
library(tidyverse)

df<-import(here("Data/clean_data.csv"))

df %>%
  filter(subject == "rla", grade != "all_grades", grade != "", !is.na(pctprof), !is.na(grade)) %>%
  mutate(grade = factor(grade, levels = rev(unique(grade)))) %>%
  mutate(grade = droplevels(grade)) %>%
  group_by(grade, group) %>%
  summarise(avg_pctprof = mean(pctprof, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(y = grade, x = avg_pctprof, color = group)) +
  geom_segment(aes(x = 0, xend = avg_pctprof, yend = grade), size = 1.2) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(round(avg_pctprof, 1), "%")),  
            hjust = 1,  
            vjust = -1.5,   
            size = 3.5,    
            fontface = "bold",
            show.legend = FALSE) +
  scale_x_continuous(limits = c(0, 100)) +
  scale_y_discrete(labels = function(x) gsub("^0", "", x)) +
  labs(
    title = "Reading Proficiency Percentage by Grade and Economic Status",
    y = "Grade",
    x = "Average Proficiency (%)", 
    subtitle = "Economically Disadvantaged Student Perform Worse in Reading Proficiency Across Grades"
  ) +
  scale_color_viridis_d(
    labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students"),
    begin = 0.1,
    end = 0.5
  ) +
  guides(color = guide_legend(title = NULL, 
                              reverse = TRUE)) +  # Swapping the order of the legend labels
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "left",
    legend.title.position = "top",
    legend.spacing.x = unit(5, "lines"),  # Space between legend items
    legend.spacing.y = unit(1, "lines"),  # Vertical space between legend and plot
    legend.margin = margin(t = -5)  # Adds space between the legend and the plot
  )

```


::: {style="text-align: center;"}
This is our final graph, chosen for its accessibility and ability to convey our message. The main takeaway is that while only about half of students in all districts are proficient in reading, this proficiency percentage is even lower when only examining economically disadvantaged (ECD) students. Also, there is variation in the gap between ECD students and all students from grade to grade, but it does not seem particularly significant. While 3rd graders are separated by about 12 percentage points, 8th graders are separated by about 13 percentage points. It is possible that the discrepancy increases as grade level increases, but this would require further statistical testing. We added subtitles to both final graphs (reading and math) to ensure our message is clear. 
:::


```{r, fig.width=12, fig.height=8}
grade_levels <- rev(sort(unique(df$grade[df$grade != "all_grades" & df$grade != "" & !is.na(df$grade)])))

df %>%
  filter(subject %in% c("rla", "math"), 
         grade != "all_grades", 
         grade != "", 
         !is.na(pctprof), 
         !is.na(grade)) %>%
  mutate(
    grade = factor(grade, levels = grade_levels),  # Use externally defined grade_levels
    subject = factor(subject, levels = c("math", "rla"), 
                     labels = c("Mathematics", "Reading (RLA)")),
    group = factor(group, levels = c("ecd", "all"), 
                   labels = c("Economically Disadvantaged", "All Students")),
    grade_numeric = as.numeric(grade)  # Convert factor to numeric for positioning
  ) %>%
  group_by(subject, grade, group) %>%
  summarise(
    avg_pctprof = mean(pctprof, na.rm = TRUE),
    grade_numeric = first(grade_numeric),  # Carry over numeric grade
    .groups = "drop"
  ) %>%
  mutate(
    grade_offset = grade_numeric + ifelse(subject == "Mathematics", 0.3, -0.3)
  ) %>%
  ggplot(aes(y = grade_offset, x = avg_pctprof, color = subject, shape = group)) +
  # Add horizontal dashed lines to distinguish grade levels
  geom_hline(yintercept = seq(1.5, length(grade_levels) - 0.5, by = 1), 
             color = "grey20", linetype = "dashed") +
  geom_segment(aes(x = 0, xend = avg_pctprof, yend = grade_offset), size = 1.5) +
  geom_point(size = 4.5) +
  scale_x_continuous(limits = c(0, 100)) +
  # Set y-axis breaks at base grade positions using grade_levels
  scale_y_continuous(breaks = seq_along(grade_levels), labels = grade_levels) +
  labs(
    title = "Proficiency Percentage by Grade, Economic Status, and Subject",
    y = "Grade Level",
    x = "Average Proficiency (%)"
  ) +
  scale_color_viridis_d(begin = 0.1, end = 0.5) +
  scale_shape_manual(values = c("Economically Disadvantaged" = 17, "All Students" = 16)) +
  guides(
    color = guide_legend(title = "Subject"),
    shape = guide_legend(title = "Economic Status")
  ) +
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "left",
    legend.title.position = "top",
    legend.spacing.x = unit(3, "lines"),
    legend.spacing.y = unit(2, "lines"),
    legend.margin = margin(t = 10, b = 10),
    axis.text.y = element_text(size = 16),
    axis.text.x = element_text(size = 16),
    plot.title = element_text(size = 24, face = "bold"),
    plot.margin = margin(30, 30, 30, 30)
  ) +
  coord_cartesian(clip = "off")

```

::: {style="text-align: center;"}

Here we created a graph with both reading and math on the same chart. We have a slight preference for keeping them separate to allow for more information to be displayed (ie. percentage labels), but this was not possible without overcrowding this version of the graph. We also added dashed lines between grade levels to make the separation easier to see at a glance. 


Please see below for more iterations of this graph.
:::

<br><br>
<br><br>
<br><br>


::: {style="text-align: center;"}
As discussed in our previous post, we experimented with multiple types of graphs and color schemes before settling on a lollipop graph. However, we still needed to make adjustments to styling and themes to have the graph be optimized for our blog. We started with the first graph, using a green and orange color palette and minimal theme. 
:::

```{r, fig.show = 'hold'}

#lollipop 1
df %>%
  filter(subject == "rla", grade != "all_grades", grade != "", !is.na(pctprof), !is.na(grade)) %>%
  mutate(grade = factor(grade, levels = rev(unique(grade)))) %>%
  mutate(grade = droplevels(grade)) %>%
  group_by(grade, group) %>%
  summarise(avg_pctprof = mean(pctprof, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(y = grade, x = avg_pctprof, color = group)) +
  geom_segment(aes(x = 0, xend = avg_pctprof, yend = grade), size = 1.2) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(round(avg_pctprof, 1), "%")),  
          hjust = 1,  
          vjust = -1.5,   
          size = 3.5,    
          fontface = "bold",
          show.legend = FALSE) +
  scale_x_continuous(limits = c(0, 100)) +
  labs(
    title = "Reading Proficiency Percentage by Grade and Economic Status",
    y = "Grade",
    x = "Average Proficiency (%)",
    color = "Economic Status")+
  scale_color_manual(
    values = c("all" = "#1b9e77", "ecd" = "#d95f02"),
    labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students")
  ) +
  theme_minimal()

```

<br>

<div style="text-align: center;">

We then decided to try making the graph color blind friendly. We added the viridis color palette but ran into an issue, each viridis palette (magma, plasma, etc.) has a bright yellow color that is very washed out on a minimal theme. We tried adding a dark theme, but that was not very aesthetically pleasing and the darker numbers were difficult to read. 
</div>
<br><br>
```{r, fig.show = 'hold'}
df %>%
  filter(subject == "rla", grade != "all_grades", grade != "", !is.na(pctprof), !is.na(grade)) %>%
  mutate(grade = factor(grade, levels = rev(unique(grade)))) %>%
  mutate(grade = droplevels(grade)) %>%
  group_by(grade, group) %>%
  summarise(avg_pctprof = mean(pctprof, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(y = grade, x = avg_pctprof, color = group)) +
  geom_segment(aes(x = 0, xend = avg_pctprof, yend = grade), size = 1.2) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(round(avg_pctprof, 1), "%")),  
            hjust = 1,  
            vjust = -1.5,   
            size = 3.5,    
            fontface = "bold",
            show.legend = FALSE) +
  scale_x_continuous(limits = c(0, 100)) +
  labs(
    title = "Reading Proficiency Percentage by Grade and Economic Status",
    y = "Grade",
    x = "Average Proficiency (%)",
    color = "Economic Status"
  ) +
  scale_color_viridis_d(
    option="magma",
    labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students")
  ) +
  theme_minimal()

df %>%
  filter(subject == "rla", grade != "all_grades", grade != "", !is.na(pctprof), !is.na(grade)) %>%
  mutate(grade = factor(grade, levels = rev(unique(grade)))) %>%
  mutate(grade = droplevels(grade)) %>%
  group_by(grade, group) %>%
  summarise(avg_pctprof = mean(pctprof, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(y = grade, x = avg_pctprof, color = group)) +
  geom_segment(aes(x = 0, xend = avg_pctprof, yend = grade), size = 1.2) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(round(avg_pctprof, 1), "%")),  
            hjust = 1,  
            vjust = -1.5,   
            size = 3.5,    
            fontface = "bold",
            show.legend = FALSE) +
  scale_x_continuous(limits = c(0, 100)) +
  labs(
    title = "Reading Proficiency Percentage by Grade and Economic Status",
    y = "Grade",
    x = "Average Proficiency (%)",
    color = "Economic Status"
  ) +
  scale_color_viridis_d(
    option="magma",
    labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students")
  ) +
  theme_dark()


```

<div style="text-align: center;">
We realized we could set a beginning and end to the viridis colors, which could eliminate the bright yellow and make the graph more friendly for colorblind and non-colorblind viewers. This also allowed us to keep the theme_minimal(), which matched better with the rest of our blog. We kept playing with the beginning and end values until we were satisfied that the colors looked good while still remaining accessible. As final touches, we adjusted the legend to be horizontal and near the top. 
</div><br>
```{r}
df %>%
  filter(subject == "rla", grade != "all_grades", grade != "", !is.na(pctprof), !is.na(grade)) %>%
  mutate(grade = factor(grade, levels = rev(unique(grade)))) %>%
  mutate(grade = droplevels(grade)) %>%
  group_by(grade, group) %>%
  summarise(avg_pctprof = mean(pctprof, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(y = grade, x = avg_pctprof, color = group)) +
  geom_segment(aes(x = 0, xend = avg_pctprof, yend = grade), size = 1.2) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0(round(avg_pctprof, 1), "%")),  
            hjust = 1,  
            vjust = -1.5,   
            size = 3.5,    
            fontface = "bold",
            show.legend = FALSE) +
  scale_x_continuous(limits = c(0, 100)) +
  scale_y_discrete(labels = function(x) gsub("^0", "", x)) +
  labs(
    title = "Reading Proficiency Percentage by Grade and Economic Status",
    y = "Grade",
    x = "Average Proficiency (%)"
  ) +
  scale_color_viridis_d(
    labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students"),
    begin = 0.1,
    end = 0.5
  ) +
  guides(color = guide_legend(title = NULL, 
                              reverse = TRUE)) +  # Swapping the order of the legend labels
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "left",
    legend.title.position = "top",
    legend.spacing.x = unit(5, "lines"),  # Space between legend items
    legend.spacing.y = unit(1, "lines"),  # Vertical space between legend and plot
    legend.margin = margin(t = -5)  # Adds space between the legend and the plot
  )





```
<br><br>
<div style="text-align: center;">
Finally, we used a colorblind function to check that our graph would be accessible for different types of colorblindness. We simplified the text in the grpah to focus on the color changes, and then used the function. This satisfied our color choices, because each group was still separately identifiable, so we decided to keep our final graph as is. 
</div><br>
```{r}
library(colorblindr)
p <- df %>%
  filter(subject == "rla", grade != "all_grades", grade != "", !is.na(pctprof), !is.na(grade)) %>%
  mutate(grade = factor(grade, levels = rev(unique(grade)))) %>%
  mutate(grade = droplevels(grade)) %>%
  group_by(grade, group) %>%
  summarise(avg_pctprof = mean(pctprof, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(y = grade, x = avg_pctprof, color = group)) +
  geom_segment(aes(x = 0, xend = avg_pctprof, yend = grade), size = 0.8) +  # Smaller lines
  geom_point(size = 3) +  # Smaller points
  scale_x_continuous(limits = c(0, 100)) +
  scale_y_discrete(labels = function(x) gsub("^0", "", x)) +
  labs(
    y = "Grade",
    x = "Average Proficiency (%)"
  ) +
  scale_color_viridis_d(
    labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students"),
    begin = 0.1,
    end = 0.5
  ) +
  guides(color = guide_legend(title = NULL)) +  # Removes the legend title
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "left",
    legend.title.position = "top",
    legend.spacing.x = unit(5, "lines"),  # Space between legend items
    legend.spacing.y = unit(1, "lines"),  # Vertical space between legend and plot
    legend.margin = margin(t = -5),  # Adds space between the legend and the plot
    plot.margin = margin(t = 20, b = 20),  # Increase margins for clarity
    axis.text.x = element_text(size = 10),  # Smaller axis text
    axis.text.y = element_text(size = 10),  # Smaller axis text
    axis.title.x = element_text(size = 12),  # Smaller x-axis title
    axis.title.y = element_text(size = 12),  # Smaller y-axis title
    plot.title = element_text(size = 14)  # Smaller title
  )

# Display the plot
cvd_grid(p)




```

