---
title: "All Students"
description: |
 Prociency Scores: Compring all students to economically disadvantaged students
author:
  - name: Aden Bhagwat, Emily Mingus, Erick Njue
    url: https://adenb0220.github.io/EDLD_652_Final/
date: 2025-03-03
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(rio)
library(here)
library(tidyverse)
```

```{r}
df<-import(here("Data/clean_data.csv"))
```

To answer question 1 that examines the relationship between economic status or Social Economic Status (SES) and the proficiency scores, we first examine the relationship between the proficiency scores of all the students and those of students from economically disadvantaged students. Figure 1 shows the Math proficiency scores of all the students against those of economically disadvantaged students and Figure 2 shows the mean percent of the students in Math by their economic status. The Math proficiency for economically disadvantaged students is lower than that of all students. 

**Figure 1**

```{r}
df %>% 
  filter(subject == "math") %>% 
  filter(grade != "all_grades") %>% 
  filter(group %in% c("all", "ecd")) %>%  # Keep only "all" and "ecd" groups
  na.omit() %>%  
  ggplot(aes(x = group, y = pctprof)) +
  geom_boxplot(aes(fill = group)) +  # Color by group for clarity
  theme_minimal() +
  scale_x_discrete(labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students")) +
  labs(
    x = "Economic Status",
    y = "Percent Proficient",
    title = "Percent of Students Proficient in Math by Economic Status",
    fill = "Student Group"
  )+
  theme(legend.position = "none")
```

**Figure 2**

```{r}
df %>%
  filter(subject == "math") %>%
  filter(grade != "all_grades") %>%
  filter(group %in% c("all", "ecd")) %>%
  na.omit() %>%
  ggplot(aes(x = group, y = pctprof, fill = group)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +  # Calculate mean
  theme_minimal() +
  scale_x_discrete(labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students")) +
  labs(
    x = "Economic Status",
    y = "Mean Percent Proficient",
    title = "Mean Percent of Students Proficient in Math by Economic Status",
    fill = "Student Group")+
 theme(legend.position = "none") 
```

Figure 3 shows the Reading proficiency scores of all the students against those of economically disadvantaged students and Figure 4 shows shows the mean percent of the students in Reading by their economic status. All students perform better than economically disadvantaged students.

**Figure 3**
```{r}
df %>% 
  filter(subject == "rla") %>%  # Filter for reading proficiency
  filter(grade != "all_grades") %>% 
  filter(group %in% c("all", "ecd")) %>%  # Keep only "all" and "ecd" groups
  na.omit() %>%  
  ggplot(aes(x = group, y = pctprof)) +
  geom_boxplot(aes(fill = group)) +  # Color by group for better distinction
  theme_minimal() +
  scale_x_discrete(labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students")) +
  labs(
    x = "Economic Status",
    y = "Percent Proficient",
    title = "Percent of Students Proficient in Reading by Economic Status",
    fill = "Student Group"
  )+
  theme(legend.position = "none")

```

**Figure 4**

```{r}
df %>% 
  filter(subject == "rla") %>%  # Filter for reading proficiency
  filter(grade != "all_grades") %>% 
  filter(group %in% c("all", "ecd")) %>%  # Keep only "all" and "ecd" groups
  na.omit() %>%  
  ggplot(aes(x = group, y = pctprof, fill = group)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +  # Bar plot with mean values
  theme_minimal() +
  scale_x_discrete(labels = c("all" = "All Students", "ecd" = "Economically Disadvantaged Students")) +
  labs(
    x = "Economic Status",
    y = "Mean Percent Proficient",
    title = "Mean Percent of Students Proficient in Reading by Economic Status",
    fill = "Student Group"
  )+
  theme(legend.position = "none")
```


